---
title: "Wideband Acoustic Immittance Data Visualization"
subtitle: "Graph Replication With SQL"
date: "Nov 26, 2024"
author: "Kevin Valdivia"
bibliography: sql.bib
format: html
toc: true # Table of contents
execute: 
  warning: false
  message: false
---


### Objective

-   **Replicate [Figure 1]((https://pmc.ncbi.nlm.nih.gov/articles/PMC7093226/)) from @Voss2019** using SQL and R:
    -   Query the WAI Database to build a dataset.
    -   Pipe it into `ggplot()` to recreate Figure 1.
-   **Analyze the @Lewis2015 data by sex**:
    -   Plot frequency vs. mean absorbance by sex for this data.
    
### Data

I will use the [Wideband Acoustic Immittance (WAI) Database](https://www.science.smith.edu/wai-database/), which I will access through SQL. The following is the connection. I will also load my libraries in this first step. The libraries used in the project are the `tidyverse` and `RMariaDB`.

```{r}
library(tidyverse)
library(RMariaDB)

con_wai <- dbConnect(
  MariaDB(), host = "scidb.smith.edu",
  user = "waiuser", password = "smith_waiDB", 
  dbname = "wai"
)
Measurements <- tbl(con_wai, "Measurements")
PI_Info <- tbl(con_wai, "PI_Info")
Subjects <- tbl(con_wai, "Subjects")
```

## Exploring The Database

There are seven datasets in the database. To achieve our objectives, we will use the `Measurements`, `Subjects`, and `PI_Info` datasets.

```{sql}
#| connection: con_wai
SHOW TABLES;
```

From the `Measurements` database, I will use the `Identifier` variable to join with the `PI_Info` dataset. I will also use the variables `Frequency` and `Absorbance` for plotting. Additionally, I will analyze the variables `Ear` and `Instrument` to determine the number of ears and the types of instruments used in each study.

```{sql}
#| connection: con_wai
DESCRIBE Measurements;
```

We will use the `PI_Info` dataset by joining it with the `Measurements` dataset using the `Identifier` variable. From this dataset, we will extract the `Year` of the study and the `AuthorsShortList` to include in the legend.

```{sql}
#| connection: con_wai
DESCRIBE PI_Info;
```

# Research Replication

I will begin by replicating the graph from Voss (2020). To do this, I will join the datasets `Measurements` and `PI_Info` using the `Identifier` key variable. I will extract the variable and calculate the mean of `Absorbance` for each value of `Frequency.` Additionally, I will create a label for each study using the relevant information from the datasets. Finally, I will save the resulting dataset as replication_data.

```{sql}
#| connection: con_wai
#| output.var: "replication_data"
SELECT 
    m.Frequency,
    AVG(m.Absorbance) as mean_absorbance,
    CONCAT(pi.AuthorsShortList, " (", pi.Year, ") N=", COUNT(DISTINCT m.SubjectNumber, m.Ear), "; ", m.Instrument) AS label
FROM Measurements as m
INNER JOIN PI_Info AS pi ON m.Identifier=pi.Identifier
WHERE m.Identifier IN ( "Abur_2014", "Feeney_2017", "Groon_2015", 
                        "Lewis_2015", "Liu_2008", "Rosowski_2012",
                        "Shahnaz_2006", "Shaver_2013", "Sun_2016",
                        "Voss_1994", "Voss_2010", "Werner_2010") 
                  AND m.Frequency <= 8000 AND m.Frequency >= 200
GROUP BY m.Identifier, Instrument, Frequency;

```

I created a custom color palette to match the lines in the graphs I am replicating.

```{r}
custom_colors <- c(
  "#5C4E9E", "#387BBC", "#5FB8A9", "#99D59F", "#D9F0A7", "#ECF5AF",
  "#F7EEAF", "#F8D37E", "#EC9D57", "#E46647", "#C9384D", "#9C2045"
)

color_map <- setNames(custom_colors, unique(replication_data$label))
```

Now, I will use the `replication_data` to create a graph that replicates the original (becuase of updates in the data there will be slight differnces in the plotting). I will use the `color_map` variable to apply the custom color template.

```{r, fig.width=10, fig.height=6}
replication_data |> 
  ggplot(aes(x = Frequency, y = mean_absorbance, color = label)) +
  geom_line(size = 1) +
  scale_x_continuous(
    trans = "log10",               
    breaks = c(200, 400, 600, 800, 1000, 2000, 4000, 6000, 8000),
    labels = scales::label_number()) +
  theme_light() +
  scale_color_manual(values = color_map) +
  guides(color = guide_legend(ncol = 1)) +
  theme(
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.5, "cm"),
    legend.spacing.y = unit(0.5, "cm")) +  
  labs(
    x = "Frequency (Hz)", 
    y = "Mean Absorbance",
    title = "Mean absorbance from each publication in WAI database",
    color = "")
```

# Graphing Lewis & Neely 2015

In the `Subjects` database, we will use the `Identifier` as a key to connect the datasets. Additionally, we will use the variable `Sex` to analyze the difference in mean absorbance by sex. We will join this database with Measurments database.

```{sql}
#| connection: con_wai
DESCRIBE Subjects;
```

Now, I will join the `Subjects` dataset with the `Measurement` dataset to plot a similar graph, but focusing only on the research from `Lewis_2015` and grouped by `Sex`.

```{sql}
#| connection: con_wai
#| output.var: "absorbance_by_sex"
SELECT 
    m.Identifier,
    m.Frequency,
    AVG(m.Absorbance) as mean_absorbance,
    s.Sex
FROM Measurements as m
INNER JOIN Subjects AS s ON m.SubjectNumber=s.SubjectNumber AND 
                            m.Identifier=s.Identifier
WHERE m.Identifier IN ("Lewis_2015") AND m.Frequency <= 8000 AND m.Frequency >= 200
GROUP BY Identifier, Frequency, Sex;
```

Now, I will graph the `mean_absorbance` against `Frequency`, grouped by `Sex`, for the `Lewis_2015` research data.

```{r}
custom_palette <- c("Female" = "darkred", 
                    "Male" = "darkgreen")

absorbance_by_sex |> 
  mutate(Sex = ifelse(Sex == "Female", "Female", "Male")) |>  # There semmed to be some variability with how Male is spelled
  ggplot(aes(x = Frequency, y = mean_absorbance, color = Sex)) +
  geom_line(size = 1.25)+  
  theme_minimal() +
  scale_color_manual(values = custom_palette) +
  labs(
    x = "Frequency (Hz)", 
    y = "Mean Absorbance",
    title = "Mean absorbance by Sex For Lewis and Neely (2015)",
    color = "Sex"
  )
```

The graph follows a similar pattern to the previously replicated graph. However, males have slightly higher mean absorbance at the start, and then there is a slight increase in mean absorbance for males around 6000. Finally, at the rightmost end, females show a much higher mean absorbance.
