---
title: "Sherlock Holmes Dataset"
subtitle: "Sentiment Analysis"
date: "Nov 18, 2025"
author: "Kevin Valdivia"
format: html
toc: false # Table of contents
execute: 
  warning: false
  message: false
---

This week [tidytuesday](https://github.com/rfordatascience/tidytuesday/blob/main/data/2025/2025-11-18/readme.md) will explore the complete line-by-line text of the Sherlock Holmes stories and novels from the R package `[sherlock](https://github.com/EmilHvitfeldt/sherlock)`.

Libraries used for the analysis

```{r}
library(tidyverse)
library(tidytext)
library(tidytuesdayR)
library(rvest)
library(tools)
```

Data imported from the tidytuesday repo

```{r}
holmes <- tidytuesdayR::tt_load('2025-11-18')$holmes |> 
  mutate(book = tolower(book)) |> 
  mutate(book = str_squish(book))
```

### Data Mainpulation

Scraping from wikipedia to get the publishing year for each of the books to be used in the plot.

```{r}
holmes_wiki <- read_html("https://en.wikipedia.org/wiki/Canon_of_Sherlock_Holmes") |> 
  html_elements("p+ ol li , p+ ol a") |> 
  html_text() |> 
  as_tibble() |> 
  filter(!is.na(str_extract(value, "\\d{4}"))) |> 
  mutate(date = str_extract(value, "\\d{4}")) |> 
  mutate(title = str_extract(value, '^"?([^"(]+)"?')) |> 
  mutate(title = str_replace_all(title, '"', '')) |> 
  mutate(title = tolower(title)) |> 
  mutate(title = str_squish(title))
```

Joining the dataset of the publishing year to the dataset with the sentiment analysis.

```{r}
holmes <- holmes |> 
  left_join(holmes_wiki, by = c("book" = "title"))
```

Pulling the sentiments from tidytext.

```{r}
afinn_sentiments <- tidytext::get_sentiments("afinn") 
nrc_sentiments <- tidytext::get_sentiments("nrc") |> 
  filter(sentiment != "positive", sentiment != "negative")
```

Joining the text with the holmes dataset.

```{r}
sent_data <- holmes |> 
  tidytext::unnest_tokens(words, text) |> 
  anti_join(stop_words, by = c("words" = "word")) |> 
  left_join(nrc_sentiments, by = c("words" = "word"), relationship = "many-to-many") |> 
  group_by(date, sentiment) |> 
  summarize(n = n()) |> 
  filter(!is.na(sentiment)) |> 
  mutate(prop = n / sum(n)) |> 
  mutate(date = as.numeric(date))
```

### Data Viz

Plot showing the changes in the sentiment of the books overtime.

```{r, fig.width=10, fig.height=5}
sent_data |> 
  ggplot(aes(x = date, y = prop)) + 
  geom_line(aes(color = sentiment), size = 1.2) + 
  geom_point(shape = 21, fill = "white", aes(color = sentiment), size = 3, alpha = 0.8) + 
  scale_color_brewer(palette = "Dark2", 
                     labels = function(x) toTitleCase(x)) +
  labs(title = "Proportion of Each Sentiment Over Time",
       subtitle = 
         "Tracking sentiment changes in the Sherlock Holmes novels by published year",
       y = "Proportion", x = "Year") + 
  
  theme_minimal(base_size = 15) + 
  theme(legend.title = element_blank(),
        plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(size = 12),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title = element_text(size = 14))

```
