---
title: "Brazilian Companies"
subtitle: "Distribution of capital stock and owner/partner qualification label "
date: "Jan 27, 2025"
author: "Kevin Valdivia"
format: html
toc: false # Table of contents
execute: 
  warning: false
  message: false
---

This weeks [TidyTuesday dataset](https://github.com/rfordatascience/tidytuesday/blob/main/data/2026/2026-01-27/readme.md) was "curated from Brazil’s open CNPJ (Cadastro Nacional da Pessoa Jurídica) records published by the Brazilian Ministry of Finance / Receita Federal on the national open-data portal (dados.gov.br)."

Libraries that will be used in the creation of the graph.

```{r}
library(tidyverse)
library(scales)
```

## Data Manipulation

Importing the companies dataset from the TidyTuesday repo.

```{r}
companies <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-01-27/companies.csv')
```

Data manipulation to put the companies with the quantiles that each owner qualification has.

```{r}
quantile_data <- companies |> 
  
  mutate(owner_qualification = case_when(
      owner_qualification == "Director / Officer" ~ "Director",
      owner_qualification == "President / Chair" ~ "President",
      owner_qualification == "Judicial Administrator (Court-appointed)" ~ "Judicial",
      owner_qualification == "Beneficial Owner (individual) resident or domiciled in Brazil" ~ "Beneficial",
      owner_qualification == "Entrepreneur / Business Owner" ~ "Entrepreneur",
      owner_qualification == "Administrator / Manager" ~ "Administrator",
      owner_qualification == "Liquidator" ~ "Liquidator",
      owner_qualification == "Attorney-in-fact / Legal Representative (Power of Attorney)" ~ "Attorney",
      owner_qualification == "Managing Partner / Partner-Administrator" ~ "Partner",
      owner_qualification == "Ostensible Partner (Managing partner in a silent partnership)" ~ "Partner",
      owner_qualification == "Executor / Estate Administrator" ~ "Executor",
      owner_qualification == "Sole Owner of an Individual Real Estate Company" ~ "Owner",
      TRUE ~ owner_qualification
    )) |> 
  
  filter(owner_qualification != "Intervenor / Court-appointed Administrator") |> 
  group_by(owner_qualification) |> 
  summarize(q25 = quantile(capital_stock, p = 0.25),
            q50 = quantile(capital_stock, p = 0.50),
            q75 = quantile(capital_stock, p = 0.75),
            q90 = quantile(capital_stock, p = 0.90),
            q99 = quantile(capital_stock, p = 0.99),
            n = n()) |> 
  
  mutate(q99 = q99 - q90,
         q90 = q90 - q75,
         q75 = q75 - q50,
         q50 = q50 - q25) |> 
  
  pivot_longer(cols = q25:q99,
               names_to = "percentile",
               values_to = "quantile") |> 
  
  arrange(desc(quantile)) |> 
  
  mutate(
    percentile = forcats::fct_rev(percentile),
    owner_qualification = forcats::fct_reorder(
      owner_qualification, 
      quantile, 
      .desc = TRUE))
```

## Map Creation

Palette that takes some colors form the Brazilian map.

```{r}
brazil_palette <- c("#009739", "#84CC16", "#FFDF00", "#002776", "#4DA6FF")
```

Map that shows the differences in the capital stock quantiles between the owner qualifications.

```{r}
quantile_data |> 
  ggplot(aes(owner_qualification, 
             quantile, 
             fill = percentile)) +
  geom_bar(stat = "identity", color = "black") +
  scale_y_continuous(
    trans = "log10",
    labels = label_number(scale_cut = cut_short_scale()))+
  coord_polar() +
  theme_minimal() +
  scale_fill_manual(values = brazil_palette) +
  labs(x = "", y = "", fill = "",
       title = "Capital Stock Quantiles by Owner Qualification Label")
```