---
title: "Winter Olympics Schedule"
subtitle: "Milan Cortina"
date: "Feburary 10, 2026"
author: "Kevin Valdivia"
format: html
  # html:
  #   code-fold: true
toc: false # Table of contents
execute: 
  warning: false
  message: false
---

The data for this week comes from the Milan Winter Olympics competition schedule. The data was compiled in this [github repo](https://github.com/chendaniely/olympics-2026).

Libraries used in this analysis
```{r}
library(tidyverse)
library(lubridate)
```


## Data Cleaning
Importing the data from the TidyTuesday repository.
```{r}
raw_schedule <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-02-10/schedule.csv') 
```

Data cleaning to create the categories of ice and snow events and creating new day and hour variables.
```{r}
schedule <- raw_schedule |> 
  select(discipline_code, discipline_name, event_code, 
         event_description, start_datetime_utc, end_datetime_utc,
         is_medal_event, venue_name, location_name, session_code, 
         start_time) |> 
  distinct() |> 
  mutate(week = lubridate::week(start_datetime_utc),
         day = lubridate::yday(start_datetime_utc)) |> 
  mutate(surface = case_when(
      discipline_name == "Bobsleigh" ~ "ice",
      discipline_name == "Curling" ~ "ice",
      discipline_name == "Figure Skating" ~ "ice",
      discipline_name == "Ice Hockey" ~ "ice",
      discipline_name == "Luge" ~ "ice",
      discipline_name == "Short Track Speed Skating" ~ "ice",
      discipline_name == "Skeleton" ~ "ice",
      discipline_name == "Speed Skating" ~ "ice",
      TRUE ~ "snow"
    ),
    start_hour = hour(start_time),
    day = wday(start_datetime_utc, week_start = 1, label = TRUE)
    )
```

## Data Viz
Vizualtion of the frequency of the events over the course of a day faceted by types of the events and days of the week.
```{r, fig.height=6}
schedule |> 
  
  mutate(surface = ifelse(surface == "ice", "Ice", "Snow")) |> 
  group_by(start_hour, surface, day) |> 
  summarize(n = n()) |> 
  
  ggplot(aes(x = start_hour, 
             y = n, 
             fill = surface)) +
  geom_col(position = "identity") +
  facet_grid(day~surface) +
  theme_linedraw() +
  scale_fill_manual(values = c("Ice"  = "lightblue", 
                               "Snow" = "darkblue"))+ 
  theme(legend.position = "none") +
  labs(x = "Hour of Day", y = "", 
       title = "Frequency of Milan Winter Olympics Events",
       subtitle = "Vertically faceted by days of the week")
```







