---
title: "Golem Grad Tortoises"
subtitle: "Patterns & X-Axis"
date: "March 3, 2026"
author: "Kevin Valdivia"
format: html
  # html:
  #   code-fold: true
toc: false # Table of contents
execute: 
  warning: false
  message: false
---

This weeks tidytuesday dataset comes from the paper ["Sex Ratio Bias Triggers Demographic Suicide in a Dense Tortoise Population"](https://onlinelibrary.wiley.com/doi/10.1111/ele.70296). The goal of this project was to make an intresting looking plot and use ggpatterns and work on makeing a y-axis level with two levels.

## Libraries
The libraries that were used
```{r}
library(tidyverse)
library(ggpattern)
```



## Data 
Importing the data from the tidytuesday repo.
```{r}

tortoise <- read_csv(
  'https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2026/2026-03-03/tortoise_body_condition_cleaned.csv'
  )

```

## Plot
Plotting to practice using patterns and sublayered x axis labels.
```{r}
tortoise |> 
  group_by(sex, locality) |> 
  mutate(ave_condition = mean(body_condition_index),
         sex = ifelse(sex == "f", "F", "M")) |> 
  
  ggplot() + 
  geom_violin_pattern(aes(y = body_condition_index, x = sex,
                  fill = locality, pattern = sex)) +
  geom_boxplot(aes(y = body_condition_index, x = sex), 
               width = 0.3, alpha = 0.3, color = "white",
               outlier.shape = NA) +
  
  scale_pattern_manual(values = c("M" = "none",
                                  "F" = "stripe"
                        #"stripe"
                        #"crosshatch"
                        #"circle"
                        #"pch" 
                        #"plasma"
                        #"gradient"
                        #“none”
  )) +
  labs(x = "",
       y = "Body Condition",
       title = "Body Condition by Sex and Locality",
       pattern = "Sex", fill = "Locality"
       ) +
  scale_fill_manual(values = c("#2F4F2F", 
                               "#6B8E23", 
                               "#A3B86C")) + 
  guides(fill = guide_legend(override.aes = 
                               list(pattern = "none"))) +
  facet_grid(~locality, switch = "x") +
  
  theme_minimal() 

```




