---
title: "Biodiverisity in West Coast National Parks"
subtitle: "Creating a Waffle Plot of Unique Species Spotted"
date: "Oct 8, 2024"
author: "Kevin Valdivia"
format: html
toc: false # Table of contents
execute: 
  warning: false
  message: false
---

### Libraries Used

Waffle was used to create the waffle graph.

```{r}
library(tidyverse)
library(waffle)
```

### Data importation

This Tidytuesday dataset was curated by [f.hull](https://github.com/frankiethull) and comes from NPSpecies - [The National Park Service biodiversity database](https://irma.nps.gov/NPSpecies/). This data was accessed on September 2nd, 2024.

```{r}
species <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-10-08/most_visited_nps_species_data.csv')
```

```{r}
species |> 
  distinct(ParkName)
```

We see the Great Smoky Mountains (GRSM) has a much higher amount of biodiveristy than any other park even when examining known present species.

```{r}
species |> 
  group_by(ParkCode, Occurrence) |> 
  summarize(biodiv = n()) |> 
  ggplot(aes(ParkCode, biodiv, color = Occurrence)) +
  geom_point(size = 2) +
  #scale_y_log10() +
  coord_flip() +
  labs(x = "Unqiue Species",
       y = "Park Code",
       color = "Occurence\nClassification",
       title = "Unique Species At Popular National Parks\nby the Species Occurence Classification")+
  theme_minimal()
```

# Creating a Waffle plot

First I create a dataset called wc_species that has the west coast national parks and filters for

```{r}
wc_species <- species |> 
  count(ParkCode, CategoryName, ParkName) |> 
  filter(n>=10 & ParkCode %in% c("JOTR", "YOSE", "OLYM")) |> 
  mutate(n = round(n/10))
```

### Creating a color pallete

Here I identify the unique biological categories in west coast naitonal parks.

```{r}
unique(wc_species$CategoryName)
```

I use ChatGPT to create a color scheme based of off htese biological categories that will be easy to see in a waffle chart.

```{r}
custom_palette <- c(
  "Bacteria" = "#FF6347",            
  "Bird" = "#1E90FF",                
  "Fungi" = "#A0522D",               
  "Insect" = "#FFD700",              
  "Mammal" = "#B22222",              
  "Other Non-vertebrates" = "#7CFC00",
  "Reptile" = "#9ACD32",             
  "Spider/Scorpion" = "#8A2BE2",     
  "Vascular Plant" = "#4E7C31",      
  "Amphibian" = "#DAA520",           
  "Fish" = "#4682B4"                 
)
```

## Plotting

I plot a waffle chart that shows the biodiversity of biological groups amoung the three west coast naitonal parks in the dataset.

```{r}
wc_species |> 
  mutate(ParkName = str_extract(ParkName, ".*(?= National Park)")) |> 
  ggplot(aes(fill = CategoryName, values = n)) +
  geom_waffle(color = "white", size=0.25, n_rows = 10, flip = TRUE) +
  facet_wrap(~ParkName, strip.position = "bottom") +
  scale_x_discrete() +
  scale_y_continuous(label = function(x) x * 100, 
                    expand = c(0,0)) +
  coord_equal() +
  theme_minimal() +
theme(
    panel.grid = element_blank(), 
    axis.ticks.y = element_line()
    #strip.text = element_text(face = "bold")  # Make facet labels bold
  ) +
  guides(fill = guide_legend(reverse = TRUE)) +
  scale_fill_manual(values = custom_palette) +
  labs(fill = "Biological Group",
       title = "Biodiversity in West Coast National Parks",
       subtitle = "Each Square Represents 10 Species by Biological Category")
```
