---
title: "ISO Country Codes"
subtitle: "Plotly lollipop graph of when country codes where discontinued"
date: "Nov 12, 2024"
author: "Kevin Valdivia"
format: html
toc: false # Table of contents
execute: 
  warning: false
  message: false
---
**The dataset this week comes from the {[ISOcodes](https://cran.r-project.org/web/packages/ISOcodes/index.html)} R package.  My analysis will focus on the `former_countries` dataset  which is of formerly used names of countries from ISO 3166-3.**

For this TidyTuesday, I’m creating an interactive lollipop chart to show when different countries ceased to be countries using the former_countries dataset.

### Libraries Used
```{r}
library(tidyverse)
library(lubridate)
library(plotly)
```

### Data Importation
```{r}
former_countries <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-11-12/former_countries.csv')
```

## Data Preparation
```{r}
former <- former_countries |> 
  mutate(date = case_when(
    nchar(date_withdrawn) == 4 ~ ymd(paste0(date_withdrawn, "-01-01")),
    TRUE ~ ymd(date_withdrawn))) |> 
  mutate(yval = sample(5:20, nrow(former_countries), replace = TRUE)) |> 
  arrange(date) |> 
  mutate(decade = as.character((((year(date) %% 100)) %/% 10) * 10)) |> 
  mutate(decade = case_when(decade == "0" ~ "00",
                            TRUE ~ decade)) |> 
  mutate(decade = factor(decade, levels = c("70", "80", "90", "00", "10", "20"))) |> 
  mutate(name = paste0(year(date), ": ", name))

```

## Data Vizualization
```{r}
# Define custom colors for each decade
decade_colors <- c("70" = "#1f3a6c", 
                   "80" = "#6a1b9a", 
                   "90" = "#e53946", 
                   "00" = "#f57c00", 
                   "10" = "#43a047", 
                   "20" = "#00897b")

# Reorder `name` by `date`
former <- former |> 
  mutate(alpha_3 = fct_reorder(alpha_3, date))

# Create the plot
p <- former |> 
  ggplot(aes(x = alpha_3, y = date)) +
  geom_segment(aes(xend = alpha_3, yend = min(date), color = decade)) +
  geom_point(aes(text = name, color = decade), size = 4) +
  scale_color_manual(values = decade_colors) +
  theme_minimal() +
  labs(title = "Interactive Lollipop Plot", 
       x = "Country Code", 
       y = "Date",
       color = "Decade") +
  coord_flip()

# Make it a plotly
ggplotly(p, tooltip = "text")

```
