---
title: "Creating a Word Map With Custom Shapes"
subtitle: "Dataset is About The Words Used in Shakespeare's Romeo & Juliet"
date: "Sep 17, 2024"
author: "Kevin Valdivia"
format: html
toc: false # Table of contents
execute: 
  warning: false
  message: false
---

```{r}
library(tidyverse)
library(gganimate)
library(maps)
```


```{r}
country_results_df <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-09-24/country_results_df.csv')
individual_results_df <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-09-24/individual_results_df.csv')
#timeline_df <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-09-24/timeline_df.csv')
```

In 2024, there were 609 people competing from 109 different countries.

```{r}
indiv_country <- individual_results_df |> 
  mutate(country2 = ifelse(str_detect(country, "C[0-9]"), "Russia", country)) 

indiv_country |> 
  filter(year == 2024) |> 
  summarise(distinct_countries = n_distinct(country2), n_countries = n())
```


```{r}
indiv_country <- indiv_country |> 
  group_by(country2, year) |> 
  arrange(individual_rank) |> 
  slice_head(n = 1) |> 
  arrange(desc(year)) |> 
  select(year, country2, individual_rank, award) |> 
  mutate(country3 = case_when(
    country2 == "United States of America" ~ "USA",
    country2 == "People's Republic of China" ~ "China",
    country2 == "Türkiye" ~ "Turkey",
    country2 == "Islamic Republic of Iran" ~ "Iran",
    country2 == "Republic of Korea" ~ "South Korea",
    country2 == "Islamic Republic of Iran" ~ "Iran",
    country2 == "Turkish Republic of Northern Cyprus" ~ "Cyprus",
    country2 == "Democratic People's Republic of Korea" ~ "North Korea",
    country2 == "United Kingdom" ~ "UK",
    TRUE ~ country2
  )) |> 
  mutate(award = as.factor(case_when(
         str_detect(award, "Gold") ~ "Gold",
         str_detect(award, "Silver") ~ "Silver",
         str_detect(award, "Bronze") ~ "Bronze",
         TRUE ~ NA))) |> 
  mutate(award = fct_relevel(award, c("Gold", "Silver", "Bronze")))

indiv_country
```


```{r}
world_data <- map_data("world")
world_data |>
 summarize(n_distinct(region))
plot_imo <- world_data |> 
  left_join(indiv_country, by = c("region" = "country3")) 
```



library(gganimate)
my_breaks <- c(1, 10, 50, 200, 600)

plot_imo |> 
  mutate(year = as.factor(year)) |> 
  #filter(year == 1979 | !is.na(individual_rank)) |> 
  filter(!is.na(individual_rank)) |> 
ggplot(aes(x = long, y = lat, group = group)) + 
  geom_polygon(data = world_data, 
               aes(x = long, y = lat, group = group),
               color = "black", fill = "white") + 
  geom_polygon(aes(fill = individual_rank, color = award)) + 
  scale_fill_gradient(name = "best individual rank", 
                      trans = "log",
                      low = "#71797E",
                      high = "#E6E6E6",
                      breaks = my_breaks, labels = my_breaks) + 
  scale_color_manual(values = c("#FFD700", "#C0C0C0", "#CE8946"),
                     na.value="black",
                     breaks = c("Gold", "Silver", "Bronze")) +
  gganimate::transition_states(year, 
                               state_length = 5,
                               transition_length = 10) +
  labs(title = 'Math Olympiad top scorer in {closest_state}')


