---
title: "Economic Diversity In Elite Schools"
subtitle: "Interactive Data Vizualization (select to highlight)"
date: "Sep 10, 2024"
author: "Kevin Valdivia"
format: html
toc: false # Table of contents
execute: 
  warning: false
  message: false
---

The data explores economic diversity and student outcomes. The dataset this week comes from [Opportunity Insights](https://opportunityinsights.org/data/) and is described in its [codebook.](https://opportunityinsights.org/wp-content/uploads/2023/07/CollegeAdmissions_Codebook.pdf)

In this tidytuesday we will make an interactive plotly graph, that has select to highlight capability, between the variables `attend` and `par_income_bin` faceted by whether the school is public or private.

### Loading relevant libraries

```{r}
library(tidyverse)
library(plotly)
library(crosstalk) # For SharedData
```

### Importation of the data

```{r}
college_admissions <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-09-10/college_admissions.csv')
```

## Setting up interactive graph

::: {.callout-note title="Plotly & Crosstalk::SharedData"}
Reminder: the SharedData function hass to be used in this way for the plotly to work.
:::

```{r}
# Some Data prep so that public and private categroization looks good on the graph
college_admissions <- college_admissions |>
  mutate(public_name = ifelse(public == TRUE, "Public University", "Private University"))

# Indispensable setup for plotly
data_w_crosstalk <- crosstalk::SharedData$new(college_admissions, ~name)
```

## ggplot Setup

First we need to make the ggplot look how we want it to look.

```{r}
plot_setup <- data_w_crosstalk |> 
  ggplot(aes(x = par_income_bin, y = attend, color = public_name,
             text = paste0("School: ", name))) +
  geom_point() +
  geom_line() +
  facet_grid(~ public_name) +
    scale_color_manual(values = c("Public University" = "#024ebf", "Private University" = "#bf0202")) +  # Custom colors
  guides(color = "none") + # How to not make the color and public/private appear
  theme_minimal() +
  labs(x = "Parental Income Bin",
       y = "Attned Variable",
       title = "Attend Variable by Parental Income Bin for Elite Schools")
```

# Plotly

Now we can use the plotly to make the ggplot saved interactive

```{r}
plotly_setup <- plotly::ggplotly(plot_setup, tooltip = "text")
highlight(plotly_setup, on = "plotly_hover")
```

*When hovering over the lines, we can observe the relationship between parental income percentile and attendance for each of the 139 institutions in the study. Private schools generally increase at the highest income bins, while most public schools decrease for the highest incomes.*

More Information About The Variables:

-   `Attend Variable` : Test-score-reweighted absolute attendance rate: Calculated as the fraction of students attending that college among all test-takers within a parent income bin in the Pipeline Analysis Sample.

-   `Parental Income Bin Variable` : Parent household income group based on percentile in the income distribution.
