---
title: "Dialogue From Bob's Burgers"
subtitle: "Plotting Burgers In ggplot"
date: "Nov 19, 2024"
author: "Kevin Valdivia"
format: html
toc: true # Table of contents
knitr:
  opts_chunk: 
    dev: "ragg_png"
execute: 
  echo: true
  warning: false
  message: false
---
This week we’re exploring Bob’s Burgers dialogue! Instead of using the official tidytuesday rep data we will use the `[bobsburgersR](https://github.com/poncest/bobsburgersR)` data package. Thank you to [Steven Ponce](https://github.com/poncest) for the data,

### Libraries
Loading the necessary libraries. By loading the bobsburgersR package our data will also be loaded as `transcript_data` and `imdb_wikipedia_data`.
```{r}
library(tidyverse) 
library(bobsburgersR)
```

## Data Preparation

We will be making text analysis graphs realted to burger related lines, so we will first make a list of "burger keywords". Then we put it in a regex format to detect whether a line has a burger keyword.

```{r}
burger_keywords <- c(
  "patty", "burger", "cheeseburger", "double-decker", 
  "bacon burger", "slider", "smashburger", 
  "grilled bun", "toasted bun", "drive-thru", 
  "ketchup", "mustard", "mayo", "pickles", "lettuce", 
  "tomato", "onions", "cheese", "toppings", "fries"
)
burger_keywords_list <- str_c("(?i)", str_c(burger_keywords, collapse = "|"))
```

Now I will put the data in a format so we can plot how frequently burger related words where said accross seasons and episodes. I will also join the data with the imdb data to also have a rating for another plot.
```{r}
plot_data <- transcript_data |> 
  drop_na(raw_text) |> 
  mutate(word_in = str_detect(raw_text, burger_keywords_list)) |> 
  group_by(season, episode) |> 
  summarize(num_wrds = sum(word_in), .groups = "drop") |> 
  full_join(imdb_wikipedia_data, by = c("season", "episode"))
```

## Graphing

First I will facet wrap a plot of the requency of burger lines across the seasons.
```{r}
#| fig-width: 10
#| fig-height: 5

season_labels <- setNames(
  paste("Season", 1:14),
  as.character(1:14))

plot_data |> 
  ggplot(aes(y = num_wrds, x = episode)) + 
  geom_text(
    aes(label = "hamburger"), 
    size = 2.5, color = "#D2691E", 
    family = "Font Awesome 5 Free Solid") +
  facet_wrap(
    ~season, 
    labeller = as_labeller(season_labels)) + 
  theme_minimal() +
  labs(
    subtitle = "Number of times lines had burger-related words",
    title = "Dialogue from Bob's Burgers",
    x = "Episode Number",
    y = "Number of Lines per Episode")
```
Now I will make a scatterplot to view the relationship between "burger lines" and the imdb rating.
```{r}
plot_data |> 
  ggplot(aes(num_wrds, rating)) +
  geom_text(
    aes(label = "hamburger"), 
    size = 2.5, color = "#D2691E", 
    family = "Font Awesome 5 Free Solid") +  
  geom_smooth(color = "darkgreen",
              se = FALSE) +
  theme_minimal() +
  labs(title = "Bob's Burgers: Number of Burger Related Lines Against Episode Rating",
       subtitle = "We see very little relationship as seen by the green trend line",
       x = "Number of Burger Related Lines",
       y = "Episode Rating")
```


