---
title: "National Highways Traffic Flow"
subtitle: ""
date: "Dec 3, 2024"
author: "Kevin Valdivia"
format: html
toc: false # Table of contents
execute: 
  warning: false
  message: false
---

Thank you to [Nicola Rennie](https://github.com/nrennie) for curating this week's dataset. The data can be accessed via the [National Highways API](https://webtris.nationalhighways.co.uk/api/swagger/ui/index). It has the vehicle volume from 4 highway cameras along an English highway, the A64.

### Libraries

```{r}
library(tidyverse)
```

## Data Importation and Prep

There are a few [English Bank Holidays](https://www.gov.uk/bank-holidays) in our dataset specifically May 3rd (Early May bank holiday) and May 31st (Spring bank holiday) in 2021. COVID restrictions are [lifted](https://www.gov.uk/government/news/further-easing-of-covid-restrictions-confirmed-for-17-may) in May 17 of 2021 in England and we will also take this into account. We will encode these details to observe them in our graphs.

```{r}
A64_traffic <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-12-03/A64_traffic.csv') |> 
  mutate(dow = wday(`Report Date`, label = TRUE)) |> 
  mutate(holiday = ifelse(`Report Date` == ymd("2021-05-03") | `Report Date` == ymd("2021-05-31"), TRUE, FALSE)) |> 
  mutate(status = ifelse(`Report Date` < "2021-05-17", "covid_strict", "covid_free")) # |> 
  # mutate(label = case_when(
  #   SiteId == 6867 ~ "Norton-on-Derwent",
  #   SiteId == 7035 ~ "B1249 & A1039",
  #   SiteId == 6867 ~ "York (north)",
  #   SiteId == 6867 ~ "Filey"
  # ))
```

## Graphing

We will compare vehicle volume during Monday holidays against other Mondays of the month as well as the presence of COVID restrictions.

```{r}
A64_traffic |> 
  filter(dow == "Mon") |> 
  filter(SiteId == 6867) |> 
  mutate(holiday = ifelse(holiday == TRUE, "Yes", "No")) |> 
  ggplot(aes(x = `Time Period Ending`, y = `Total Volume`, color = holiday, group = `Report Date`, linetype = status)) + 
  theme_minimal() +
  scale_color_manual(
    values = c("Yes" = "darkgreen", "No" = "red"),
    name = "Holiday"
  ) +
  scale_linetype_manual(
    values = c("covid_free" = "solid", "covid_strict" = "dashed"),
    labels = c("covid_free" = "Lifted", "covid_strict" = "Active"),
    name = "Covid Restrictions") +
  geom_line() +
  labs(title = "Vehicle Volumed In The A64 Highway (siteID 6867)",
       subtitle = "The first Mon with COVID restrictions lifted and a holiday \nhas an increased vehicle volume",
       x = "Hour of Day",
       y = "Vehicle Volume",
       color = "Holiday")
```

Now I will facet by day of the week, vehicle type, camera on the highway, and covid restrictions.

```{r}
#| fig-width: 13
#| fig-height: 9
#| fig-cap: More driving after COVID restrictions were lifted can be observed in the graph. Note that the orange lines represent the number of cars and should be interpreted using the left y-axis, while the blue lines represent trucks and should be interpreted using the right y-axis.

A64_traffic |> 
  ggplot(aes(
    x = hour(`Time Period Ending`), 
    group = `Report Date`, 
    linetype = status
  )) + 
  geom_smooth(aes(y = `0 - 520 cm`, color = "Cars"), se = FALSE) + 
  geom_smooth(aes(y = `661 - 1160 cm` * 10, color = "Trucks"), se = FALSE) + 
  facet_grid(SiteId ~ dow) + 
  scale_y_continuous(
    # Features of the first axis
    name = "Number of Cars",
    # Add a second axis and specify its features
    sec.axis = sec_axis(~./10, name = "Number of Small Trucks")
  ) + 
  scale_color_manual(
    values = c("Cars" = "darkorange", "Trucks" = "blue"),
    name = "Vehicle Type"
  ) +
  scale_linetype_manual(
    values = c("covid_free" = "solid", "covid_strict" = "dashed"),
    labels = c("covid_free" = "Lifted", "covid_strict" = "Active"),
    name = "Covid Restrictions"
  ) +
  theme(
    axis.title.y = element_text(color = "darkorange", size = 13),
    axis.title.y.right = element_text(color = "blue", size = 13)
  ) +
  labs(title = "Number of Vehicles Captured By Cameras On A64 Highway",
       subtitle = "Columns represents days of the week while rows represent highways cameras. \nEach line is a different week and the line type distiguishes the type of covid restrictions in place.",
       x = "Hour of Day") +
  theme_minimal()


```
